\subsection{Multi-Token Prediction (MTP)}
\label{subsec:mtp}
Multi-Token Prediction (MTP) has emerged as a highly effective technique for improving both the 
\emph{accuracy} and the \emph{inference efficiency} of large language models. Prior work—
including DeepSeek-V3~\citep{deepseekai2025deepseekv3technicalreport} and the original MTP 
formulation~\citep{gloeckle2024better}—shows that predicting multiple future tokens provides richer 
training signals and encourages models to plan several steps ahead. These auxiliary predictions also 
serve naturally as draft tokens for speculative decoding~\citep{leviathan2023fast}, enabling 
substantial end-to-end acceleration without requiring a separate draft model.

In Nemotron~3, integrating MTP leads to consistent gains across validation loss and a broad range of downstream benchmarks, including general knowledge, code generation, common-sense understanding, reading comprehension, and math. On an ablation study using the 8B active parameters transformer MoE model base model, MTP improves performance by roughly 2.4\% on average across benchmarks (see Table~\ref{tab:moe_mtp_results}). These improvements reflect MTP’s ability to provide denser supervision and enhance the model’s multi-step reasoning capabilities. From a systems perspective, MTP introduces minimal additional FLOPs and integrates seamlessly into our training workflow, delivering substantial speculative-decoding benefits~\citep{gloeckle2024better} while maintaining high overall efficiency.


\begin{table}[h!]
\centering
\begin{tabular}{l|cc}
\toprule
\textbf{Task} & 
\textbf{Baseline (8B MoE Base)} &
\textbf{Baseline + MTP} \\
\midrule

\rowcolor{black!5}
\multicolumn{3}{l}{\textbf{General Knowledge}} \\
MMLU (5-shot, acc)        & 70.06 & \textbf{71.26} \\
MMLU-Pro (5-shot, CoT EM) & 45.05 & \textbf{47.84} \\
\midrule

\rowcolor{black!5}
\multicolumn{3}{l}{\textbf{Code}} \\
MBPP-Sanitized (3-shot) & 65.58 & \textbf{66.89} \\
% Note: Code avg actually decreases (59.21 → 56.76), so NOT included
% If you want only improvements, REMOVE this row
\midrule

\rowcolor{black!5}
\multicolumn{3}{l}{\textbf{Commonsense Understanding}} \\
ARC-Challenge (25-shot, acc\_norm) & 86.43 & \textbf{88.05} \\
WinoGrande (0-shot, acc)        & 74.59 & \textbf{75.45} \\
\midrule
\rowcolor{black!5}
\multicolumn{3}{l}{\textbf{Reading Comprehension}} \\
RACE (0-shot, acc)              & 84.02 & \textbf{85.36} \\
\midrule
\rowcolor{black!5}
\multicolumn{3}{l}{\textbf{Math}} \\
% MATH and math500 both decrease, so excluded
GSM8K (8-shot, acc)   & 82.49 & \textbf{84.46} \\
\bottomrule
\end{tabular}

\caption{
Accuracy scores with and without MTP on a simple 8B active parameter transformer MoE base model trained on 1T tokens. We observe improvements in accuracy on multiple tasks spanning different categories.
}
\label{tab:moe_mtp_results}
\end{table}


A major practical benefit of MTP in Nemotron~3 is its strong synergy with speculative decoding. 
The predictions produced by MTP exhibit high agreement with the base model, enabling fast, 
low-latency generation — particularly beneficial in batch-size--1 and long-form generation scenarios.  
We designed a lightweight MTP module that achieves around 97\% acceptance on the first two predicted tokens in an ablation study on an 8B active MoE model. Overall, MTP enriches the training 
signal, enhances the model’s ability to anticipate multiple future steps, provides high-quality 
speculative-decoding predictions, and accelerates text generation~\citep{gloeckle2024better}. 

%This stability supports more aggressive speculative 
%decoding strategies and contributes directly to faster, more fluid model responses. % , and maintains reasonably high acceptance for deeper draft sequences evlauated on a base model



% MTP is a key enabler of Nemotron~3’s accuracy and performance. It

