\section{Pre-training}

In this section, we describe the construction of our pretraining data, the details of our pretraining approach, and present experimental results from evaluating the base models on standard benchmarks.

\label{sec:pre}

\subsection{Pre-training Data}
Compared with Qwen2.5 \citep{qwen2.5}, we have significantly expanded the scale and diversity of our training data. Specifically, we collected twice as many pre-training tokens—covering three times more languages. All Qwen3 models are trained on a large and diverse dataset consisting of \textbf{119 languages and dialects}, with a total of \textbf{36 trillion tokens}. This dataset includes high-quality content in various domains such as coding, STEM (Science, Technology, Engineering, and Mathematics), reasoning tasks, books, multilingual texts, and synthetic data.

To further expand the pre-training data corpus, we first employ the Qwen2.5-VL model \citep{qwen2.5vl} to perform text recognition on a large volume of PDF-like documents. The recognized text is then refined using the Qwen2.5 model \citep{qwen2.5}, which helps improve its quality. Through this two-step process, we are able to obtain an additional set of high-quality text tokens, amounting to trillions in total. 
Besides, we employ Qwen2.5 \citep{qwen2.5}, Qwen2.5-Math \citep{qwen2.5math}, and Qwen2.5-Coder \citep{qwen2.5coder} models to synthesize trillions of text tokens in different formats, including textbooks, question-answering, instructions, and code snippets, covering dozens of domains. 
Finally, we further expand the pre-training corpus by incorporating additional multilingual data and introducing more languages. Compared to the pre-training data used in Qwen2.5, the number of supported languages has been significantly increased from 29 to 119, enhancing the model's linguistic coverage and cross-lingual capabilities.

We have developed a multilingual data annotation system designed to enhance both the quality and diversity of training data.
This system has been applied to our large-scale pre-training datasets, annotating over 30 trillion tokens across multiple dimensions such as educational value, fields, domains, and safety. These detailed annotations support more effective data filtering and combination. 
Unlike previous studies \citep{doremi,doge,regmix} that optimize the data mixture at the data source or domain level, our method optimizes the data mixture at the instance-level through extensive ablation experiments on small proxy models with the fine-grained data labels.




\subsection{Pre-training Stage}
The Qwen3 models are pre-trained through a three-stage process:

\begin{enumerate}[label=(\arabic*)]
    \item \textbf{General Stage (S1)}: At the first pre-training stage, all Qwen3 models are trained on over 30 trillion tokens using a sequence length of 4,096 tokens. At this stage, the models have been fully pre-trained on language proficiency and general world knowledge, with training data covering 119 languages and dialects.  
    \item \textbf{Reasoning Stage (S2)}: To further improve the reasoning ability, we optimize the pre-training corpus of this stage by increasing the proportion of STEM, coding, reasoning, and synthetic data. The models are further pre-trained with about 5T higher-quality tokens at a sequence length of 4,096 tokens. We also accelerate the learning rate decay during this stage.
    \item \textbf{Long Context Stage}: In the final pre-training stage, we collect high-quality long context corpora to extend the context length of Qwen3 models. All models are pre-trained on hundreds of billions of tokens with a sequence length of 32,768 tokens. The long context corpus includes 75\% of text between 16,384 to 32,768 tokens in length, and 25\% of text between 4,096 to 16,384 in length. Following Qwen2.5 \citep{qwen2.5}, we increase the base frequency of RoPE from 10,000 to 1,000,000 using the ABF technique \citep{ropeabf}. Meanwhile, we introduce YARN \citep{yarn} and Dual Chunk Attention~(DCA, \citealp{chunkllama}) to achieve a four-fold increase in sequence length capacity during inference.
\end{enumerate}

Similar to Qwen2.5 \citep{qwen2.5}, we develop scaling laws for optimal hyper-parameters (e.g., learning rate scheduler, and batch size) predictions based on three pre-training stages mentioned above. Through extensive experiments, we systematically study the relationship between model architecture, training data, training stage, and optimal training hyper-parameters. Finally, we set the predicted optimal learning rate and batch size strategy for each dense or MoE model.




\subsection{Pre-training Evaluation}

We conduct comprehensive evaluations of the base language models of the Qwen3 series. 
The evaluation of base models mainly focuses on their performance in general knowledge, reasoning, mathematics, scientific knowledge, coding, and multilingual capabilities. The evaluation datasets for pre-trained base models include 15 benchmarks:


%
\begin{itemize}

\item \textbf{General Tasks}: MMLU \citep{mmlu} (5-shot), MMLU-Pro \citep{mmlupro} (5-shot, CoT), MMLU-redux \citep{mmluredux} (5-shot), BBH \citep{bbh} (3-shot, CoT), SuperGPQA \citep{supergpqa}(5-shot, CoT).

\item \textbf{Math \& STEM Tasks}: GPQA \citep{gpqa} (5-shot, CoT), GSM8K \citep{gsm8k} (4-shot, CoT), MATH \citep{math} (4-shot, CoT).

\item \textbf{Coding Tasks}: EvalPlus \citep{evalplus} (0-shot) (Average of HumanEval \citep{humaneval}, MBPP \citep{mbpp}, Humaneval+, MBPP+) \citep{evalplus}, MultiPL-E \citep{multiple} (0-shot) (Python, C++, JAVA, PHP, TypeScript, C\#, Bash, JavaScript), MBPP-3shot \citep{mbpp}, CRUX-O of CRUXEval (1-shot) \citep{gu2024cruxeval}.

\item \textbf{Multilingual Tasks}: MGSM \citep{mgsm} (8-shot, CoT), MMMLU \citep{mmmlu} (5-shot), INCLUDE \citep{romanou2024includeevaluatingmultilinguallanguage} (5-shot).

\end{itemize}


For the base model baselines, we compare the Qwen3 series base models with the Qwen2.5 base models \citep{qwen2.5} and other leading open-source base models, including DeepSeek-V3 Base \citep{deepseekv3}, Gemma-3 \citep{gemma3}, Llama-3 \citep{llama3}, and Llama-4 \citep{llama4} series base models, in terms of scale of parameters. All models are evaluated using the same evaluation pipeline and the widely-used evaluation settings to ensure fair comparison.


\paragraph{Summary of Evaluation Results}
Based on the overall evaluation results, we highlight some key conclusions of Qwen3 base models. 
\begin{enumerate}[label=(\arabic*)]
\item Compared with the previously open-source SOTA dense and MoE base models (such as DeepSeek-V3 Base, Llama-4-Maverick Base, and Qwen2.5-72B-Base), Qwen3-235B-A22B-Base outperforms these models in most tasks with significantly fewer total parameters or activated parameters.

\item For the Qwen3 MoE base models, our experimental results indicate that: (a) Using the same pre-training data, Qwen3 MoE base models can achieve similar performance to Qwen3 dense base models with only \textbf{1/5} activated parameters. (b) Due to the improvements of the Qwen3 MoE architecture, the scale-up of the training tokens, and more advanced training strategies, the Qwen3 MoE base models can outperform the Qwen2.5 MoE base models with less than \textbf{1/2} activated parameters and fewer total parameters. (c) Even with \textbf{1/10} of the activated parameters of the Qwen2.5 dense base model, the Qwen3 MoE base model can achieve comparable performance, which brings us significant advantages in inference and training costs.

\item The overall performance of the Qwen3 dense base models is comparable to the Qwen2.5 base models at higher parameter scales. For example, Qwen3-1.7B/4B/8B/14B/32B-Base achieve comparable performance to Qwen2.5-3B/7B/14B/32B/72B-Base, respectively. Especially in STEM, coding, and reasoning benchmarks, the performance of Qwen3 dense base models even surpasses Qwen2.5 base models at higher parameter scales.
\end{enumerate}
The detailed results are as follows.

\begin{table}[tbp]
\centering
\caption{\textbf{Comparison among Qwen3-235B-A22B-Base and other representative strong open-source baselines. The highest,
the second-best scores are shown in \textbf{bold} and \underline{underlined}, respectively.}}
\label{tab:base-235A22B}
\small %
\setlength{\tabcolsep}{3pt} %
%\renewcommand{\arraystretch}{0.9} 
\begin{tabular}{@{}lccccc@{}}
\toprule
& \textbf{Qwen2.5-72B} & \textbf{Qwen2.5-Plus} & \textbf{Llama-4-Maverick} & \textbf{DeepSeek-V3} & \textbf{Qwen3-235B-A22B} \\
& \textbf{Base} & \textbf{Base} & \textbf{Base} & \textbf{Base} & \textbf{Base} \\
\midrule
Architecture & Dense & MoE & MoE & MoE & MoE\\
\# Total Params & 72B & 271B & 402B & 671B & 235B\\
\# Activated Params & 72B & 37B & 17B & 37B & 22B\\
\midrule
\multicolumn{6}{c}{\textit{General Tasks}} \\
\midrule
MMLU & 86.06 & 85.02 & 85.16 & \underline{87.19} & \textbf{87.81}\\
MMLU-Redux & 83.91 & 82.69 & 84.05 & \underline{86.14} & \textbf{87.40} \\
MMLU-Pro & 58.07 & 63.52 & \underline{63.91} & 59.84 & \textbf{68.18} \\
SuperGPQA & 36.20 & 37.18 & 40.85 & \underline{41.53} & \textbf{44.06} \\
BBH & \underline{86.30} & 85.60 & 83.62 & 86.22 & \textbf{88.87} \\
\midrule
\multicolumn{6}{c}{\textit{Math \& STEM Tasks}} \\
\midrule
GPQA & \underline{45.88} & 41.92 & 43.94 & 41.92 & \textbf{47.47} \\
GSM8K & 91.50 & \underline{91.89} & 87.72 & 87.57 & \textbf{94.39} \\
MATH & 62.12 & 62.78 & \underline{63.32} & 62.62 & \textbf{71.84} \\
\midrule
\multicolumn{6}{c}{\textit{Coding Tasks}} \\
\midrule
EvalPlus & 65.93 & 61.43 & \underline{68.38} & 63.75 & \textbf{77.60} \\
MultiPL-E & 58.70 & 62.16 & 57.28 & \underline{62.26} & \textbf{65.94} \\
MBPP & \underline{76.00} & 74.60 & 75.40 & 74.20 & \textbf{81.40} \\
CRUX-O & 66.20 & 68.50 & \underline{77.00} & 76.60 & \textbf{79.00} \\
\midrule
\multicolumn{6}{c}{\textit{Multilingual Tasks}} \\
\midrule
MGSM & 82.40 & 82.21 & 79.69 & \underline{82.68} & \textbf{83.53} \\
MMMLU & 84.40 & 83.49 & 83.09 & \underline{85.88} & \textbf{86.70} \\
INCLUDE & 69.05 & 66.97 & \underline{73.47} & \textbf{75.17} & 73.46 \\
\bottomrule
\end{tabular}
\end{table}



\paragraph{Qwen3-235B-A22B-Base}
We compare Qwen3-235B-A22B-Base to our previous similar-sized MoE Qwen2.5-Plus-Base \citep{qwen2.5} and other leading open-source base models: Llama-4-Maverick \citep{llama4}, Qwen2.5-72B-Base \citep{qwen2.5}, DeepSeek-V3 Base \citep{deepseekv3}. From the results in Table~\ref{tab:base-235A22B}, the Qwen3-235B-A22B-Base model attains the highest performance scores across most of the evaluated benchmarks. We further compare Qwen3-235B-A22B-Base with other baselines separately for the detailed analysis.
\begin{enumerate}[label=(\arabic*)]
    \item  Compared with the recently open-source model Llama-4-Maverick-Base, which has about \textbf{twice} the number of parameters, Qwen3-235B-A22B-Base still performs better on most benchmarks.
    \item  Compared with the previously state-of-the-art open-source model DeepSeek-V3-Base, Qwen3-235B-A22B-Base outperforms DeepSeek-V3-Base on 14 out of 15 evaluation benchmarks with only about \textbf{1/3} the total number of parameters and \textbf{2/3} activated parameters, demonstrating the powerful and cost-effectiveness of our models. 
    \item  Compared with our previous MoE Qwen2.5-Plus of similar size, Qwen3-235B-A22B-Base significantly outperforms it with fewer parameters and activated parameters, which shows the remarkable advantages of Qwen3 in pre-training data, training strategy, and model architecture.
    \item  Compared with our previous flagship open-source dense model Qwen2.5-72B-Base, Qwen3-235B-A22B-Base surpasses the latter in all benchmarks and uses fewer than \textbf{1/3} of the activated parameters. Meanwhile, due to the advantage of the model architecture, the inference costs and training costs on each trillion tokens of Qwen3-235B-A22B-Base are much cheaper than those of Qwen2.5-72B-Base.
\end{enumerate}


\begin{table}[tbp]
\centering
\caption{\textbf{Comparison among Qwen3-32B-Base and other strong open-source baselines. The highest and second-best scores are shown in \textbf{bold} and \underline{underlined}, respectively.}}
\label{tab:base-32B}
\small
\begin{tabular}{@{}lccccc@{}}
\toprule
 & \textbf{Qwen2.5-32B} & \textbf{Qwen2.5-72B} & \textbf{Gemma-3-27B} & \textbf{Llama-4-Scout} & \textbf{Qwen3-32B} \\
 & \textbf{Base} & \textbf{Base} & \textbf{Base} & \textbf{Base}& \textbf{Base} \\
\midrule
Architecture & Dense & Dense & Dense & MoE & Dense\\
\# Total Params & 32B & 72B & 27B & 109B & 32B\\
\# Activated Params & 32B & 72B & 27B & 17B & 32B\\
\midrule
\multicolumn{6}{c}{\textit{General Tasks}} \\
\midrule
MMLU & 83.32 & \textbf{86.06} & 78.69 & 78.27 & \underline{83.61} \\
MMLU-Redux & 81.97 & \textbf{83.91} & 76.53 & 71.09 & \underline{83.41} \\
MMLU-Pro & 55.10 & \underline{58.07} & 52.88 & 56.13 & \textbf{65.54} \\
SuperGPQA & 33.55 & \underline{36.20} & 29.87 & 26.51 & \textbf{39.78} \\
BBH & 84.48 & \underline{86.30} & 79.95 & 82.40 & \textbf{87.38} \\
\midrule
\multicolumn{6}{c}{\textit{Math \& STEM Tasks}} \\
\midrule
GPQA & \underline{47.97} & 45.88 & 26.26 & 40.40 & \textbf{49.49} \\
GSM8K & \underline{92.87} & 91.50 & 81.20 & 85.37 & \textbf{93.40} \\
MATH & 57.70 & \textbf{62.12} & 51.78 & 51.66 & \underline{61.62} \\
\midrule
\multicolumn{6}{c}{\textit{Coding Tasks}} \\
\midrule
EvalPlus & \underline{66.25} & 65.93 & 55.78 & 59.90 & \textbf{72.05} \\
MultiPL-E & 58.30 & \underline{58.70} & 45.03 & 47.38 & \textbf{67.06} \\
MBPP & 73.60 & \underline{76.00} & 68.40 & 68.60 & \textbf{78.20} \\
CRUX-O & \underline{67.80} & 66.20 & 60.00 & 61.90 & \textbf{72.50} \\
\midrule
\multicolumn{6}{c}{\textit{Multilingual Tasks}} \\
\midrule
MGSM & 78.12 & \underline{82.40} & 73.74 & 79.93 & \textbf{83.06} \\
MMMLU & 82.40 & \textbf{84.40} & 77.62 & 74.83 & \underline{83.83} \\
INCLUDE & 64.35 & \textbf{69.05} & \underline{68.94} & 68.09 & 67.87 \\
\bottomrule
\end{tabular}
\end{table}


\begin{table}[tbp]
\centering
\caption{\textbf{Comparison among Qwen3-14B-Base, Qwen3-30B-A3B-Base, and other strong open-source baselines. The highest and second-best scores are shown in \textbf{bold} and \underline{underlined}, respectively.}}
\label{tab:base-14B}
\adjustbox{center=\textwidth}{
\small
\setlength{\tabcolsep}{3pt} %
\begin{tabular}{@{}lcccccc@{}}
\toprule
 & \textbf{Gemma-3-12B} & \textbf{Qwen2.5-14B} & \textbf{Qwen2.5-32B} & \textbf{Qwen2.5-Turbo} & \textbf{Qwen3-14B} & \textbf{Qwen3-30B-A3B} \\
 &   \textbf{Base} & \textbf{Base} & \textbf{Base} & \textbf{Base}& \textbf{Base} & \textbf{Base}\\
\midrule
Architecture & Dense & Dense & Dense & MoE & Dense & MoE\\
\# Total Params & 12B & 14B & 32B & 42B & 14B & 30B\\
\# Activated Params & 12B & 14B & 32B & 6B & 14B & 3B\\
\midrule
\multicolumn{7}{c}{\textit{General Tasks}} \\
\midrule
MMLU & 73.87 & 79.66 & \textbf{83.32} & 79.50 & 81.05 & \underline{81.38} \\
MMLU-Redux & 70.70 & 76.64 & \textbf{81.97} & 77.11 & 79.88 & \underline{81.17} \\
MMLU-Pro & 44.91 & 51.16 & 55.10 & 55.60 & \underline{61.03} & \textbf{61.49} \\
SuperGPQA & 24.61 & 30.68 & 33.55 & 31.19 & \underline{34.27} & \textbf{35.72} \\
BBH & 74.28 & 78.18 & \textbf{84.48} & 76.10 & 81.07 & \underline{81.54} \\
\midrule
\multicolumn{7}{c}{\textit{Math \& STEM Tasks}} \\
\midrule
GPQA & 31.31 & 32.83 & \textbf{47.97} & 41.41 & 39.90 & \underline{43.94} \\
GSM8K & 78.01 & 90.22 & \textbf{92.87} & 88.32 & \underline{92.49} & 91.81 \\
MATH & 44.43 & 55.64 & 57.70 & 55.60 & \textbf{62.02} & \underline{59.04} \\
\midrule
\multicolumn{7}{c}{\textit{Coding Tasks}} \\
\midrule
EvalPlus & 52.65 & 60.70 & 66.25 & 61.23 & \textbf{72.23} & \underline{71.45} \\
MultiPL-E & 43.03 & 54.79 & 58.30 & 53.24 & \underline{61.69} & \textbf{66.53} \\
MBPP & 60.60 & 69.00 & \underline{73.60} & 67.60 & 73.40 & \textbf{74.40} \\
CRUX-O & 52.00 & 61.10 & \underline{67.80} & 60.20 & \textbf{68.60} & 67.20 \\
\midrule
\multicolumn{7}{c}{\textit{Multilingual Tasks}} \\
\midrule
MGSM & 64.35 & 74.68 & 78.12 & 70.45 & \textbf{79.20} & \underline{79.11} \\
MMMLU & 72.50 & 78.34 & \textbf{82.40} & 79.76 & 79.69 & \underline{81.46} \\
INCLUDE & 63.34 & 60.26 & 64.35 & 59.25 & \underline{64.55} & \textbf{67.00} \\
\bottomrule
\end{tabular}
}
\end{table}

\begin{table}[tbp]
\centering
\caption{\textbf{Comparison among Qwen8B-Base and other strong open-source baselines. The highest and second-best scores are shown in \textbf{bold} and \underline{underlined}, respectively.}}
\label{tab:base-8B}
\small
\begin{tabular}{@{}lcccc@{}}
\toprule
 & \textbf{Llama-3-8B} & \textbf{Qwen2.5-7B} & \textbf{Qwen2.5-14B}  & \textbf{Qwen3-8B} \\
&   \textbf{Base} & \textbf{Base} & \textbf{Base} & \textbf{Base} \\
\midrule
Architecture & Dense & Dense & Dense & Dense\\
\# Total Params & 8B & 7B & 14B & 8B \\
\# Activated Params & 8B & 7B & 14B & 8B \\
\midrule
\multicolumn{5}{c}{\textit{General Tasks}} \\
\midrule
MMLU & 66.60 & 74.16 & \textbf{79.66} & \underline{76.89}  \\
MMLU-Redux & 61.59 & 71.06 & \textbf{76.64} & \underline{76.17}  \\
MMLU-Pro & 35.36 & 45.00 &  \underline{51.16} & \textbf{56.73}  \\
SuperGPQA & 20.54 & 26.34 & \underline{30.68} & \textbf{31.64}  \\
BBH & 57.70 & 70.40 & \underline{78.18} & \textbf{78.40} \\
\midrule
\multicolumn{5}{c}{\textit{Math \& STEM Tasks}} \\
\midrule
GPQA & 25.80 & \underline{36.36} & 32.83 & \textbf{44.44} \\
GSM8K & 55.30 & 85.36 & \textbf{90.22} & \underline{89.84} \\
MATH & 20.50 & 49.80 & \underline{55.64} & \textbf{60.80}  \\
\midrule
\multicolumn{5}{c}{\textit{Coding Tasks}} \\
\midrule
EvalPlus & 44.13 & \underline{62.18} & 60.70 & \textbf{67.65} \\
MultiPL-E & 31.45 & 50.73 & \underline{54.79} & \textbf{58.75}  \\
MBPP & 48.40 & 63.40 & \underline{69.00} & \textbf{69.80} \\
CRUX-O & 36.80 & 48.50 & \underline{61.10} & \textbf{62.00}  \\
\midrule
\multicolumn{5}{c}{\textit{Multilingual Tasks}} \\
\midrule
MGSM & 38.92 & 63.60 & \underline{74.68} & \textbf{76.02}  \\
MMMLU & 59.65 & 71.34 & \textbf{78.34} & \underline{75.72} \\
IINCLUDE & 44.94 & 53.98 & \textbf{60.26} & \underline{59.40} \\
\bottomrule
\end{tabular}
%\vspace{2mm}
\end{table}


\begin{table}[tbp]
\centering
\caption{\textbf{Comparison among Qwen3-4B-Base and other strong open-source baselines. The highest and second-best scores are shown in \textbf{bold} and \underline{underlined}, respectively.}}
\label{tab:base-4B}
\small %
%\setlength{\tabcolsep}{3pt} %
%\renewcommand{\arraystretch}{0.9} 
\begin{tabular}{@{}lcccc@{}}
\toprule
 & \textbf{Gemma-3-4B} & \textbf{Qwen2.5-3B} & \textbf{Qwen2.5-7B} & \textbf{Qwen3-4B} \\
 & \textbf{Base} & \textbf{Base}& \textbf{Base}& \textbf{Base} \\
\midrule
Architecture & Dense & Dense & Dense & Dense\\
\# Total Params & 4B & 3B & 7B & 4B \\
\# Activated Params & 4B & 3B & 7B & 4B \\
\midrule
\multicolumn{5}{c}{\textit{General Tasks}} \\
\midrule
MMLU & 59.51 & 65.62 & \textbf{74.16} & \underline{72.99} \\
MMLU-Redux & 56.91 & 63.68 & \underline{71.06} & \textbf{72.79} \\
MMLU-Pro & 29.23 & 34.61 & \underline{45.00} & \textbf{50.58} \\
SuperGPQA & 17.68 & 20.31 & \underline{26.34} & \textbf{28.43} \\
BBH & 51.70 & 56.30 & \underline{70.40} & \textbf{72.59} \\
\midrule
\multicolumn{5}{c}{\textit{Math \& STEM Tasks}} \\
\midrule
GPQA & 24.24 & 26.26 & \underline{36.36} & \textbf{36.87} \\
GSM8K & 43.97 & 79.08 & \underline{85.36} & \textbf{87.79} \\
MATH & 26.10 & 42.64 & \underline{49.80} & \textbf{54.10} \\
\midrule
\multicolumn{5}{c}{\textit{Coding Tasks}} \\
\midrule
EvalPlus & 43.23 & 46.28 & \underline{62.18} & \textbf{63.53} \\
MultiPL-E & 28.06 & 39.65 & \underline{50.73} & \textbf{53.13} \\
MBPP & 46.40 & 54.60 & \underline{63.40} & \textbf{67.00} \\
CRUX-O & 34.00 & 36.50 & \underline{48.50} & \textbf{55.00} \\
\midrule
\multicolumn{5}{c}{\textit{Multilingual Tasks}} \\
\midrule
MGSM & 33.11 & 47.53 & \underline{63.60} & \textbf{67.74} \\
MMMLU & 59.62 & 65.55 & \underline{71.34} & \textbf{71.42} \\
INCLUDE & 49.06 & 45.90 & \underline{53.98} & \textbf{56.29} \\
\bottomrule
\end{tabular}
\end{table}



\begin{table}[tbp]
\centering
\caption{\textbf{Comparison among Qwen3-1.7B-Base, Qwen3-0.6B-Base, and other strong open-source baselines. The highest and second-best scores are shown in \textbf{bold} and \underline{underlined}, respectively.}}
\label{tab:base-1B}
\small %
%\setlength{\tabcolsep}{3pt} %
%\renewcommand{\arraystretch}{0.9} 
\begin{tabular}{@{}lccccc@{}}
\toprule
 & \textbf{Qwen2.5-0.5B} & \textbf{Qwen3-0.6B} & \textbf{Gemma-3-1B} & \textbf{Qwen2.5-1.5B} & \textbf{Qwen3-1.7B} \\
 & \textbf{Base} & \textbf{Base} & \textbf{Base} & \textbf{Base} & \textbf{Base} \\
\midrule
Architecture & Dense & Dense & Dense & Dense & Dense\\
\# Total Params & 0.5B & 0.6B & 1B & 1.5B & 1.7B \\
\# Activated Params & 0.5B & 0.6B & 1B & 1.5B & 1.7B \\
\midrule
\multicolumn{6}{c}{\textit{General Tasks}} \\
\midrule
MMLU & 47.50 & 52.81 & 26.26 & \underline{60.90} & \textbf{62.63} \\
MMLU-Redux & 45.10 & 51.26 & 25.99 & \underline{58.46} & \textbf{61.66} \\
MMLU-Pro & 15.69 & 24.74 & 9.72 & \underline{28.53} & \textbf{36.76} \\
SuperGPQA & 11.30 & 15.03 & 7.19 & \underline{17.64} & \textbf{20.92} \\
BBH & 20.30 & 41.47 & 28.13 & \underline{45.10} & \textbf{54.47} \\
\midrule
\multicolumn{6}{c}{\textit{Math \& STEM Tasks}} \\
\midrule
GPQA & 24.75 & \underline{26.77} & 24.75 & 24.24 & \textbf{28.28} \\
GSM8K & 41.62 & 59.59 & 2.20 & \underline{68.54} & \textbf{75.44} \\
MATH & 19.48 & 32.44 & 3.66 & \underline{35.00} & \textbf{43.50} \\
\midrule
\multicolumn{6}{c}{\textit{Coding Tasks}} \\
\midrule
EvalPlus & 31.85 & 36.23 & 8.98 & \underline{44.80} & \textbf{52.70} \\
MultiPL-E & 18.70 & 24.58 & 5.15 & \underline{33.10} & \textbf{42.71} \\
MBPP & 29.80 & 36.60 & 9.20 & \underline{43.60} & \textbf{55.40} \\
CRUX-O & 12.10 & 27.00 & 3.80 & \underline{29.60} & \textbf{36.40} \\
\midrule
\multicolumn{6}{c}{\textit{Multilingual Tasks}} \\
\midrule
MGSM & 12.07 & 30.99 & 1.74 & \underline{32.82} & \textbf{50.71} \\
MMMLU & 31.53 & 50.16 & 26.57 & \underline{60.27} & \textbf{63.27} \\
INCLUDE & 24.74 & 34.26 & 25.62 & \underline{39.55} & \textbf{45.57} \\
\bottomrule
\end{tabular}
\end{table}



\paragraph{Qwen3-32B-Base}
Qwen3-32B-Base is our largest dense model among the Qwen3 series. We compare it to the baselines of similar sizes, including Gemma-3-27B \citep{gemma3} and Qwen2.5-32B \citep{qwen2.5}. In addition, we introduce two strong baselines: the recently open-source MoE model Llama-4-Scout, which has three times the parameters of Qwen3-32B-Base but half the activated parameters; and our previous flagship open-source dense model Qwen2.5-72B-Base, which has more than twice the number of parameters compared to Qwen3-32B-Base.
The results are shown in Table~\ref{tab:base-32B}, which support three key conclusions:
\begin{enumerate}[label=(\arabic*)]
\item Compared with the similar-sized models, Qwen3-32B-Base outperforms Qwen2.5-32B-Base and Gemma-3-27B Base on most benchmarks. Notably, Qwen3-32B-Base achieves 65.54 on MMLU-Pro and 39.78 on SuperGPQA, significantly outperforming its predecessor Qwen2.5-32B-Base. In addition, Qwen3-32B-Base achieves significantly higher encoding benchmark scores than all baseline models.
\item Surprisingly, we find that Qwen3-32B-Base achieves competitive results compared to Qwen2.5-72B-Base. Although Qwen3-32B-Base has less than half the number of parameters of Qwen2.5-72B-Base, it outperforms Qwen2.5-72B-Base in 10 of the 15 evaluation benchmarks. On coding, mathematics, and reasoning benchmarks, Qwen3-32B-Base has remarkable advantages.
\item Compared to Llama-4-Scout-Base, Qwen3-32B-Base significantly outperforms it on all 15 benchmarks, with only one-third of the number of parameters of Llama-4-Scout-Base, but twice the number of activated parameters.
\end{enumerate}

\paragraph{Qwen3-14B-Base \& Qwen3-30B-A3B-Base}
The evaluation of the Qwen3-14B-Base and Qwen3-30B-A3B-Base is compared against baselines of similar sizes, including Gemma-3-12B Base, Qwen2.5-14B Base. Similarly, we also introduce two strong baselines: (1) Qwen2.5-Turbo \citep{qwen2.5}, which has 42B parameters and 6B activated parameters. Note that its activated parameters are twice those of Qwen3-30B-A3B-Base. (2) Qwen2.5-32B-Base, which has 11 times the activated parameters of Qwen3-30B-A3B and more than twice that of Qwen3-14B.
The results are shown in Table~\ref{tab:base-14B}, where we can draw the following conclusions.
\begin{enumerate}[label=(\arabic*)]
\item Compared with the similar-sized models, Qwen3-14B-Base significantly performs better than Qwen2.5-14B-Base and Gemma-3-12B-Base on all 15 benchmarks. 
\item Similarly, Qwen3-14B-Base also achieves very competitive results compared to Qwen2.5-32B-Base with less than half of the parameters.
\item With only 1/5 activated non-embedding parameters, Qwen3-30B-A3B significantly outperforms Qwen2.5-14B-Base on all tasks, and achieves comparable performance to Qwen3-14B-Base and Qwen2.5-32B-Base, which brings us significant advantages in inference and training costs.
\end{enumerate}



\paragraph{Qwen3-8B / 4B / 1.7B / 0.6B-Base}
For edge-side models, we take similar-sized Qwen2.5, Llama-3, and Gemma-3 base models as the baselines. The results can be seen in Table~\ref{tab:base-8B}, Table~\ref{tab:base-4B}, and Table~\ref{tab:base-1B}. All Qwen3 8B / 4B / 1.7B / 0.6B-Base models continue to maintain strong performance across nearly all benchmarks. Notably, Qwen3-8B / 4B / 1.7B-Base models even outperform larger size Qwen2.5-14B / 7B / 3B Base models on over half of the benchmarks, especially on STEM-related and coding benchmarks, reflecting the significant improvement of the Qwen3 models.

