\section{Introduction}
\label{sec: introduction}


A broad goal of deep learning is to learn meaningful patterns from raw data, automatically extracting features and building abstractions in an end-to-end fashion.
However, fixed-vocabulary \textbf{tokenization}, the process of compressing raw text into predefined chunks through algorithms such as byte-pair encoding (BPE)~\citep{BPE,SentencePiece}, remains a pervasive handcrafted preprocessing step in modern language models (LMs)~\citep{Llama3, GPT3}.
Tokenization comes with a host of well-documented drawbacks,
from poor character-level understanding to lack of meaning and interpretability to degraded performance on complex languages and modalities~\citep{petrov2023language, ahia2023all, belinkov2017synthetic, Adv-BERT, ByT5, Canine}.
\footnote{Many other edge cases have been discussed in informal online discourse rather than papers; we defer to Andrej Karpathy's \href{https://x.com/karpathy/status/1657949234535211009}{lectures} and \href{https://x.com/karpathy/status/1759996551378940395}{tweets}.}
Replacing the tokenization--LM--detokenization pipeline with a single end-to-end model would also adhere better to the spirit of deep learning, ideally scaling more powerfully with data and parameters (c.f. \emph{the bitter lesson})~\citep{sutton2019bitter,peric2025bitter}.
However, tokenization remains an indispensable component of language models and other sequential data for its ability to compress and shorten sequences; as of yet, no \emph{end-to-end} tokenizer-free model has matched the performance of tokenizer-based language models when matched for computational budget.

A line of recent works has turned to overcoming tokenization in autoregressive sequence models, which requires addressing a series of difficult technical challenges:
\footnote{An extended related work can be found in \cref{sec: background}, which is summarized in \cref{tab: related}.}
\begin{itemize}
    \item Direct byte-level language modeling with isotropic architectures\footnote{Non-hierarchical models comprised of repeated blocks, such as the standard Transformer~\citep{Transformer}.} can be improved with efficient sequence models such as MambaByte~\citep{MambaByte}, 
    but still incur prohibitive computational costs while underperforming tokenized models in compute-matched settings.
    \item To improve efficiency, hierarchical architectures such as Hourglass Transformer~\citep{HourglassTransformer} and MegaByte~\citep{MegaByte} use small byte-level models to compress raw inputs into subsampled sequences, which are then processed with a more powerful standard language model. However, simple pooling strategies such as compressing every $k$ inputs are not data-dependent, and perform poorly on modalities with variable information rates such as language.
    \item SpaceByte~\citep{SpaceByte} and Byte Latent Transformer~\citep{BLT} introduce data-dependent chunking strategies such as delimiter- or entropy-based heuristics. These heuristics, however, rely on auxiliary \emph{external} boundary predictors, and are therefore modality-specific and not fully end-to-end.
    \item
    Although jointly trainable boundary predictors are the ideal solution, they require optimizing discrete selection operations without supervision, which is fundamentally a challenging problem.
    Consequently, existing end-to-end approaches~\citep{DPT} exhibit training instabilities that preclude scaling beyond small models or nesting multi-level hierarchies.
\end{itemize}
Fundamentally, creating a tokenizer-free architecture requires incorporating the data chunking process directly into the model, while overcoming challenges in efficiency, learnability, and stability at scale.

\subsubsection*{Dynamic Chunking: End-to-end Sequence Modeling Without Tokenization}

In this work, we introduce an end-to-end \textbf{hierarchical network (H-Net)} that compresses raw data through a recursive, data-dependent \textbf{dynamic chunking (DC)} process (\cref{fig: main}).
H-Nets match the efficiency of tokenized pipelines while substantially improving modeling ability,
by replacing handcrafted heuristics with content-aware and context-dependent segmentation learned from data.

\paragraph{Hierarchical Processing.}
The H-Net adopts the hierarchical architecture from prior work~\citep{SaShiMi,HourglassTransformer,SpaceByte}, resembling an autoregressive U-Net~\citep{U-Net}: (i) raw data is processed by a small \textbf{\encoder{}}, (ii) then downsampled and passed through a \textbf{\main{}} operating on compressed chunks, (iii) and finally upsampled before being passed through a \textbf{\decoder{}} operating on the original resolution.
This modularity creates a natural processing hierarchy where outer stages capture fine-grained patterns while inner stages operate on coarse representations akin to traditional tokens.
Crucially, while the main network contains the bulk of parameters and can be any standard architecture designed for operating on tokenized language---such as a Transformer~\citep{Transformer} or state space model (SSM)~\citep{Mamba}---we show that the encoder and decoder networks are strongly improved by using SSMs, which have an inductive bias for compression~\citep{gu2025tradeoffs}.



\paragraph{Dynamic Chunking.}
H-Net's core is a novel dynamic chunking (DC) mechanism which interfaces between the main network and the encoder/decoder networks, learning how to segment data while using standard differentiable optimization.
DC is composed of two complementary new techniques: (i) a \textbf{routing module} which predicts boundaries between adjacent elements through a similarity score (ii) and a \textbf{smoothing module} which interpolates representations using the router's outputs, attenuating the effect of uncertain boundaries and significantly improving learnability.
By combining these with a new auxiliary loss function that targets desired downsampling ratios, and modern techniques for gradient-based learning of discrete choices~\citep{MoE,STE},
DC lets an H-Net learn how to compress data in a fully end-to-end fashion.

\paragraph{Signal Propagation.}
We introduce several architectural and training techniques to improve stability and scalability during end-to-end optimization. 
These include: (i) carefully placing projections and normalization layers to balance signal propagation between interacting sub-networks, and (ii) adjusting optimization parameters for each layer based on its dimensionality and effective batch size, which changes between stages of the hierarchical structure.

Altogether, H-Net learns segmentation strategies \emph{optimized jointly} with the main backbone, dynamically compressing input vectors based on contextual information into meaningful chunks.
H-Net represents the first truly end-to-end, tokenizer-free language model:
with a single stage of dynamic chunking, a \textbf{\emph{byte-level H-Net} matches the perplexity and downstream performance of a strong \emph{BPE-tokenized Transformer}} at sizes exceeding 1B parameters. Empirically, the dynamic chunking module naturally compresses data to a similar resolution as BPE tokenizers (4.5-5 bytes/chunk) and qualitatively learns meaningful boundaries, all without any external supervision or heuristics.

\input{fig/main}

\vspace*{1em}

\subsubsection*{Hierarchical Chunking: From Data to Abstractions}


Beyond addressing tokenization, H-Net improves general sequence modeling across a wide range of settings.
Subword tokenization in language models is a special case of \emph{chunking}---the process of building higher-level abstractions from low-level data---and is a central component of intelligence.\footnote{\href{https://en.wikipedia.org/wiki/Chunking_(psychology)}{Chunking} is a formal concept from cognitive psychology central to human memory and cognition, and is the inspiration for this work's terminology.}
Crucially, because H-Net is fully end-to-end, \textbf{it can be iterated recursively: the main network can itself be an H-Net}.
Intuitively, more stages of chunking represent higher order meanings; just as characters can be combined into words, words can be combined into clauses, sentences, and beyond.
Iterating the hierarchy should therefore lead to even more efficient use of compute and parameters, and more effective reasoning over compressed representations.


Recursive H-Nets represent a new class of foundation model architectures that not only overcome tokenization, but discover and operate over abstractions learned from raw data, leading to higher-quality models with less pre-processing.
Iterating the 1-stage H-Net to 2 hierarchical stages further improves its capabilities and strongly outperforms all baselines, with steeper training curves and better scaling with data.
A byte-level 2-stage H-Net overtakes the perplexity of a strong tokenized Transformer after just 30B training bytes, with the gap widening throughout training, and matches the downstream evaluations of the tokenized Transformer of twice its size.

Finally, H-Nets realize the benefits of overcoming tokenization:
\begin{itemize}
    \item \textit{Robustness}.
    Without special data mixes, the pretrained H-Net is dramatically more robust to textual perturbations than the token-based Transformer, as evaluated on the noisy HellaSwag suite of benchmarks.
    \item \textit{Interpretability}.
    Qualitative visualizations of learned boundaries reveal that H-Net automatically discovers semantically coherent units without explicit supervision, validating that end-to-end learning successfully detects the structural patterns traditionally imposed through handcrafted tokenization.
    \item \textit{Other languages}.
    H-Net's improvements are even more pronounced on languages without obvious segmentation cues, including Chinese and code ($59.9 \to 66.3$ on XWinograd-zh compared to tokenized Transformer) and DNA language modeling ($3.6\times$ improved data efficiency compared to isotropic models).
\end{itemize}

We publicly release model code\footnote{\url{https://github.com/goombalab/hnet}} and pretrained checkpoints\footnote{\url{https://huggingface.co/cartesia-ai}}. 


