\chapter{Filtering, Deduplication, and Data Mixing}
\label{ch:23}

% ─── Learning Goals ───
\begin{keyinsight}
After reading this chapter, you will be able to:
\begin{enumerate}
  \item TODO: Learning goal 1
  \item TODO: Learning goal 2
  \item TODO: Learning goal 3
\end{enumerate}
\end{keyinsight}

% ─── Big Picture ───
\section{Big Picture}
\label{ch:23:sec:big_picture}

TODO: Situate this chapter in the overall narrative.

% ─── Formal Setup ───
\section{Formal Setup and Notation}
\label{ch:23:sec:setup}

TODO: Define notation and state assumptions.

% ─── Main Ideas ───
\section{Main Ideas}
\label{ch:23:sec:main}

TODO: Core technical content.

% ─── Worked Example ───
\section{Worked Example}
\label{ch:23:sec:example}

TODO: Complete worked example.

% ─── Systems Consequences ───
\section{Systems and Implementation Consequences}
\label{ch:23:sec:systems}

TODO: Implementation and performance implications.

% ─── Neuroscience Analogy ───
\begin{analogybox}{TODO: Title}
  \paragraph{Where the analogy helps.}
  TODO

  \paragraph{Where the analogy breaks.}
  TODO

  \paragraph{What intuition to keep.}
  TODO
\end{analogybox}

% ─── Misconceptions ───
\section{Common Misconceptions}
\label{ch:23:sec:misconceptions}

\begin{enumerate}
  \item \textbf{Misconception 1.} TODO: State and correct.
  \item \textbf{Misconception 2.} TODO: State and correct.
  \item \textbf{Misconception 3.} TODO: State and correct.
\end{enumerate}

% ─── Exercises ───
\section{Exercises}
\label{ch:23:sec:exercises}

\subsection*{Warmup}
\begin{enumerate}
  \item TODO
\end{enumerate}

\subsection*{Derivation}
\begin{enumerate}
  \item TODO
\end{enumerate}

\subsection*{Implementation}
\begin{enumerate}
  \item TODO
\end{enumerate}

\subsection*{Open-Ended}
\begin{enumerate}
  \item TODO
\end{enumerate}

% ─── Further Reading ───
\section{Further Reading}
\label{ch:23:sec:further}

\begin{itemize}
  \item TODO: Annotated reference 1
  \item TODO: Annotated reference 2
  \item TODO: Annotated reference 3
\end{itemize}
