% This must be in the first 5 lines to tell arXiv to use pdfLaTeX, which is strongly recommended.
\pdfoutput=1
% In particular, the hyperref package requires pdfLaTeX in order to break URLs across lines.

\documentclass[11pt]{article}

% Remove the "review" option to generate the final version.
\usepackage{acl}

% Standard package includes
\usepackage{times}
\usepackage{latexsym}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{multirow}

% For proper rendering and hyphenation of words containing Latin characters (including in bib files)
\usepackage[T1]{fontenc}
% For Vietnamese characters
% \usepackage[T5]{fontenc}
% See https://www.latex-project.org/help/documentation/encguide.pdf for other character sets

% This assumes your files are encoded as UTF8
\usepackage[utf8]{inputenc}
\usepackage{graphicx}


% This is not strictly necessary, and may be commented out,
% but it will improve the layout of the manuscript,
% and will typically save some space.
\usepackage{microtype}

% If the title and author information does not fit in the area allocated, uncomment the following
%
%\setlength\titlebox{<dim>}
%
% and set <dim> to something 5cm or larger.

\newcommand{\tablewhitespace}{\addlinespace[0.3em]}

\newcommand{\OPT}[0]{{OPT-175B}}
\newcommand{\davinci}[0]{{Davinci}}
\newcommand{\todo}[1]{{\textcolor{red}{TODO: #1}}}
\newcommand{\needcite}[0]{{(NEEDCITE)}}
\newcommand{\biggpt}[1]{{GPT-3#1}}


\title{OPT: Open Pre-trained Transformer Language Models}

\author{
{\bf Susan Zhang}\thanks{\ \ Equal contribution.
}\footnotemark[1]{\normalfont ,} 
{\bf Stephen Roller}\footnotemark[1]{\normalfont ,} 
{\bf Naman Goyal}\footnotemark[1]{\normalfont ,} 
\\
{\bf Mikel Artetxe}{\normalfont ,}
{\bf Moya Chen}{\normalfont ,}
{\bf Shuohui Chen}{\normalfont ,}
{\bf Christopher Dewan}{\normalfont ,}
{\bf Mona Diab}{\normalfont ,}
{\bf Xian Li}{\normalfont ,} \\
{\bf Xi Victoria Lin}{\normalfont ,}
{\bf Todor Mihaylov}{\normalfont ,}
{\bf Myle Ott}\thanks{\ \ Work done while at Meta AI.}\footnotemark[2]{\normalfont ,}
{\bf Sam Shleifer}\footnotemark[2]{\normalfont ,} 
{\bf Kurt Shuster}{\normalfont ,} 
{\bf Daniel Simig}{\normalfont ,} \\
{\bf Punit Singh Koura}{\normalfont ,}
{\bf Anjali Sridhar}{\normalfont ,}
{\bf Tianlu Wang}{\normalfont ,}
{\bf Luke Zettlemoyer}{\normalfont } \\
\tablewhitespace
Meta AI\\
\tablewhitespace
\texttt{\{susanz,roller,naman\}@fb.com}
}
\begin{document}
\maketitle
\begin{abstract}
Large language models, which are often trained for hundreds of thousands of compute days, have shown remarkable capabilities for zero- and few-shot learning. Given their computational cost, these models are difficult to replicate without significant capital.  For the few that are available through APIs, no access is granted to the full model weights, making them difficult to study. We present Open Pre-trained Transformers (OPT), a suite of decoder-only pre-trained transformers ranging from 125M to 175B parameters, which we aim to fully and responsibly share with interested researchers. We show that \OPT{} is comparable to \biggpt{},\footnote{Following \citet{brown2020gpt3}, we use \biggpt{} to  refer to both the 175B model and the smaller scale models as well.} while requiring only 1/7th the carbon footprint to develop.  We are also releasing our logbook detailing the infrastructure challenges we faced, along with code for experimenting with all of the released models.
\end{abstract}


\input{sections/introduction}
\input{sections/method}
\input{sections/results}
\input{sections/rai_evals}
\input{sections/limitations}
\input{sections/release}
\input{sections/related_work}
\input{sections/conclusion}

\section*{Acknowledgements}
We would like to thank Scott Jeschonek, Giri Anantharaman, Diego Sarina, Joaquin Colombo, Chris Bray, Stephen Roylance, Kalyan Saladi, Shubho Sengupta, and Brian O'Horo for helping to remove infrastructure blockers along the way; Percy Liang, Rishi Bommasani, and Emily Dinan for discussions on responsible release practices; Carole-Jean Wu for discussions on sustainability and carbon footprint considerations; Srini Iyer, Ramakanth Pasunuru, and Shruti Bhosale for previous contributions to evaluations; Benjamin Lefaudeux, Geeta Chauhan, Natalia Gimelshein, Horace He, and Sam Gross for discussions on performance improvement work; Emily Dinan, Carole-Jean Wu, Daniel McKinnon, and Mark Tygert for feedback on this draft; Antoine Bordes, Joelle Pineau, Mary Williamson, Necip Fazil Ayan, Armand Joulin, Sergey Edunov, Melanie Kambadur, Zornitsa Kozareva, Ves Stoyanov, Vitaliy Liptchinsky, Rahul Iyer, Jing Xu, Jason Weston, and many others for supporting this project internally.

\bibliography{custom,anthology}
\bibliographystyle{acl_natbib}

\clearpage
\appendix
\onecolumn

\input{sections/appendix}
\input{sections/contributions}
\input{sections/datasheet}
\input{sections/model_card}
\clearpage
\input{sections/sample_output}

\end{document}
