\normalsize
% We introduce the Nemotron 3 family of models: Nano, Super, and Ultra. Nano, the smallest model, offers better accuracy than comparable models while being extremely cost-efficient for inference. Super is optimized for collaborating agents and high-volume tasks like IT ticket automation. Ultra, the largest offering, provides state-of-the-art accuracy and reasoning abilities. The Nemotron~3 family of models offer exceptional agentic, reasoning, and chat capabilities, achieving high accuracy on academic benchmarks. Nemotron 3 models use Mixture-of-Experts hybrid Mamba-Transformer architecture to achieve best-in-class inference throughput and a context length of up to 1M tokens. Super and Ultra are trained with NVFP4 and use LatentMoE, a novel approach to improve model quality. The 2 larger models will also ship with MTP layers for faster text generation. We share all the key technologies used to train these models. We will release Nemotron 3 Nano and its technical report alongside this whitepaper. Super and Ultra model releases will follow in the upcoming months. We will openly release the model weights, recipes, and most of the training data.

% Oleksii's version:
We introduce the Nemotron 3 family of models—Nano, Super, and Ultra. These models deliver strong agentic, reasoning, and conversational capabilities. The Nemotron 3 family uses a Mixture-of-Experts hybrid Mamba–Transformer architecture to provide best-in-class throughput and context lengths of up to 1M tokens. Super and Ultra models are trained with NVFP4 and incorporate LatentMoE, a novel approach that improves model quality. The two larger models also include MTP layers for faster text generation. All Nemotron 3 models are post-trained using multi-environment reinforcement learning enabling reasoning, multi-step tool use, and support granular reasoning budget control.

Nano, the smallest model, outperforms comparable models in accuracy while remaining extremely cost-efficient for inference. Super is optimized for collaborative agents and high-volume workloads such as IT ticket automation. Ultra, the largest model, provides state-of-the-art accuracy and reasoning performance. Nano is released together with its technical report and this white paper, while Super and Ultra will follow in the coming months. We will openly release the model weights, pre- and post-training software, recipes, and all data for which we hold redistribution rights.
