\begin{abstract}

We present the design, implementation and engineering experience in building and
deploying \sysname, a production system for training large language models
(LLMs) at the scale of more than 10,000 GPUs. Training LLMs at this scale brings
unprecedented challenges to training efficiency and stability. We take a
full-stack approach that co-designs the algorithmic and system components across
model block and optimizer design, computation and
communication overlapping, operator optimization, data pipeline, and network performance tuning.
Maintaining high efficiency throughout the training process (i.e.,
stability) is an important consideration in production given the long extent of
LLM training jobs. Many hard stability issues only emerge at large scale, and
\revision{in-depth observability} is the key to address them. We develop a set of diagnosis
tools to monitor system components and events deep in the stack, identify root
causes, and derive effective techniques to achieve fault tolerance and mitigate
stragglers. \sysname achieves 55.2\% Model FLOPs Utilization (MFU) when training
a 175B LLM model on 12,288 GPUs, improving the MFU by 1.34$\times$ compared to
Megatron-LM. We share our operational experience in identifying and fixing
failures and stragglers. We hope by articulating the problems and sharing our
experience from a systems perspective, this work can inspire future LLM systems
research.

\iffalse
Large Language Models (LLMs) have shown great potential in a wide range of
downstream applications. Although the model architecture and training techniques
have been extensively studied, less light is shed on the training details of
LLMs on system level, especially at a large scale.

This work presents our experience in training a multi-billion parameter
densely-activated LLM to converge at an unprecedented scale of more than ten
thousand GPUs. The scale of infrastructure brings a series of challenges which
affects the training efficiency and stability. We propose a series of
optimizations across model design, computation and communication to achieve XX\%
Model FLOPs Utilization (MFU) \zhi{totaling xxx effective Peta-FLOPs} at this
scale. We utilize a systematic approach to accurately locate and mitigate the
straggler issues and overcame stability challenges that are typically
encountered during massive-scale training sessions. Our work paves the way for
efficiently training LLMs over ten thousands of GPUs, making it possible to
explore new frontiers in natural language understanding and generation. \zhi{may
be claiming for more rather than limited applications}
\fi
\end{abstract}
