\onecolumn

\input{sections/A_background}

\section{FLOPs Computation}
\label{sec: appendix-FLOPs_computation}
We largely follow \citet{Chinchilla} with two marginally updated computations: (1) add computations for Mamba-2~\citep{Mamba2}, and (2) modify computations in MLP blocks as we use the recent Transformer++ architecture.
Assuming that all query, key, and value share the same $\text{num\_heads}$ and $\text{head\_dim}$, we calculate the forward pass FLOPs as follows:
\begin{itemize}
    \item \textbf{Embeddings:} $2 \times \text{seq\_len} \times \text{vocab\_size} \times \text{d\_model}$
    \item \textbf{Attention:}
        \begin{itemize}
            \item \textbf{$QKV$ projections:} $2 \times 3 \times \text{seq\_len} \times \text{d\_model} \times (\text{num\_heads} \times \text{head\_dim})$
            \item \textbf{Attention Logit Calculation:} $2 \times \text{seq\_len} \times \text{seq\_len} \times (\text{num\_heads} \times \text{head\_dim})$
            \item \textbf{Attention Score Softmax:} $3 \times \text{num\_heads} \times \text{seq\_len} \times \text{seq\_len}$
            \item \textbf{Score @ Query:} $2 \times \text{seq\_len} \times \text{seq\_len} \times (\text{num\_heads} \times \text{head\_dim})$
            \item \textbf{Output projection:} $2 \times \text{seq\_len} \times (\text{num\_heads} \times \text{head\_dim}) \times \text{d\_model}$
        \end{itemize}
    \item \textbf{Mamba-2:}
        \begin{itemize}
            \item \textbf{$XZ$ projections:} $2 \times \text{seq\_len} \times \text{d\_model} \times (2 \times \text{expand} \times \text{d\_model})$
            \item \textbf{$BC \Delta t$ projections:} $2 \times \text{seq\_len} \times \text{d\_model} \times (2 \times \text{d\_state} + \text{num\_heads})$
            \item \textbf{SSD:} $2 \times 3 \times \text{seq\_len} \times (\text{expand} \times \text{d\_model}) \times \text{d\_state}$
            \item \textbf{Depthwise Convolution:} $2 \times \text{seq\_len} \times \text{d\_model} \times \text{window\_size}$
            \item \textbf{Gating:} $5 \times \text{seq\_len} \times \text{d\_model}$
            \item \textbf{Output projection:} $2 \times \text{seq\_len} \times \text{d\_model} \times \text{d\_model}$
        \end{itemize}
    \item \textbf{Gated MLP:}
        \begin{itemize}
            \item \textbf{In, Gate, Out projections:} $2 \times \text{seq\_len} \times (3 \times \text{d\_model} \times \text{ffw\_size})$
            \item \textbf{Gating:} $5 \times \text{seq\_len} \times \text{d\_model}$
        \end{itemize}
    \item \textbf{Logit Prediction Head:} $2 \times \text{seq\_len} \times \text{vocab\_size} \times \text{d\_model}$
\end{itemize}
We assume the backward pass consumes twice the FLOPs of the forward pass.

\section{Learning Rate Modulation}
\label{sec: appendix-lr_modulation}
As discussed in \cref{sec: Improved-Techniques}, we employ modulated learning rates for each stage by multiplying a scalar $\lambda^s$ to the base learning rate.
Empirically, we find that a reasonable set of multipliers (\eg{} $\lambda^0=2.0, \lambda^1=1.5, \lambda^2=1.0$) works well in general.
To provide a more systematic experimental results across different architectural configurations, we follow previous works and set learning rates to be proportionally to the (1) square root of batch size~\citep{malladi2022sdes,OLMo}, and (2) inverse square root of hidden dimension~\citep{Transformer,yang2020feature}.
Concretely, without heavy manual tuning, we define $\lambda^s$ as follows:
\begin{equation}
    \lambda^s = \sqrt{N^\text{GPT} \cdot \dfrac{\prod_{i=s}^{S} N^i}{\prod_{i=0}^{S} N^i} \cdot \dfrac{D^\text{S}}{D^s}}, \qquad N^S = 1.0
\end{equation}
where $N^\text{GPT}$ is the average number of bytes per token of training dataset, which is $4.6$ for the GPT-2 tokenizer on FineWeb-Edu.

We note that such principles for optimizing signal propagation as neural network hyperparameters change is an active area of research, and our scaling factors are just heuristics that can likely be improved.

\section{Additional Experimental Setup Details}

\subsection{Robustness Score}
\label{sec: appendix-robustness_score}

We introduce a metric called the \emph{robustness score} to measure the robustness of a model's performance to textual perturbations, defined for Hellaswag as follows: \[
    \text{robustness score} \coloneqq 100 \cdot \dfrac{\text{perturbed accuracy} - 0.25}{\max(\text{unperturbed accuracy} - 0.25, 0)}.
\]
This score measures the percentage of original (unperturbed) performance that is captured by the model in the perturbed setting. We subtract by $0.25$ as HellaSwag is multiple choice with 4 options, thus a model that scores $0.25$ in the perturbed setting should be considered to have lost all of its original capability.

\section{Additional Ablation Studies}

\input{fig/abl_pool}
\subsection{Different Downsampling Methods in the Chunking Layer}
\label{sec: Appendix-Various-Compression}
Given the dynamically determined boundaries from the boundary predictor, we explore various compression strategies in the \chunk{}.
We compare the default Downsample operation of \model{} (see \cref{sec: tokenizer}) against three alternatives (see \cref{fig: abl_pool}-left): channel-wise max/mean pooling and cross-attention, all applied to vectors within the same boundary.
Despite its simple design, the default compression in \model{} performs on-par with the other variants as demonstrated in \cref{fig: abl_pool}-right.
This shows that the sequence mixing layers in encoder are trained to implicitly compress necessary context into vectors at boundaries, without explicit compression mechanisms such as pooling or cross-attention.

\subsection{Details of Chinese and Code Experiments}

In \cref{sec: experiments-alt-language}, we analyzed the performance of \model{} (2-stage) against Transformer and \model{} (space) on Chinese and on code, finding superior scaling for \model{} (2-stage) versus the other architectures.
Here, we describe additional details from the experiment.


Besides measuring scaling behavior, we also measured final checkpoints on bits-per-byte compression ability.
We also evaluated the Chinese-language models on the Chinese split of XWinograd, a Chinese language-understanding task.
For model architecture, we primarily matched the settings from the GPT-3 \textit{XL}, including $d_{\text{model}}$ and encoder/decoder architecture for \model{} models.
However, we adjusted the number of layers in the \main{} of each model to account for slightly different compression ratios.
Specifically, the Chinese-language models used a slightly higher total training flops target than the original language models, while the code models used a lower flops target.
Full architecture details and results are also in \cref{tab: cd-supp}.

\subsection{DNA Architecture Ablations}

\label{sec: dna_supp}
\model{} (1-stage) with an \colorbox{EncDec0}{M3T1} encoder achieves $3.6\times$ the data efficiency of an isotropic architecture (\cref{fig: dna}).
As mentioned in the caption of \cref{tab: dna}, we found that an \colorbox{EncDec0}{M3T1} encoder outperformed a pure Mamba-2 \colorbox{EncDec0}{M4} encoder (\cref{tab: dna_supp}).
Putting a Transformer in the \encoder{} does not appear to be helpful for text (\cref{fig: abl_enc_dec_combination}). 
Thus, it is possible the Transformer being useful is a DNA-specific result.

Interestingly, the loss curve for the \colorbox{EncDec0}{M4} encoder with a pure Mamba-2 \main{} was more unstable.
We then also tried replacing the \colorbox{Main}{M15} in the \main{} with a \colorbox{Main}{T1M13T1} architecture,
inspired by the finding that Transformer layers are good for dealing directly with compressed input (see \cref{fig: abl_enc_dec_combination_BPE}).
The new, principled \main{} architecture improved stability greatly (\cref{fig: dna_supp}).

\input{fig/dna_supp}

\section{Distilling Token-Level Models to Byte-Level}
\label{sec: distill}

\input{fig/distill}
\input{tab/distill}

The role of the outer stages in \model{} is analagous to that of the tokenizer, embedding module, and LM head in a traditional BPE-based language model; together, these modules interconvert between raw text and an embedding space that the main model backbone can process.
Given this similarity, we investigated whether it would be possible to convert a BPE-tokenized model directly into a byte level \model{}.
To do this, we trained a 1-stage \model{} with frozen \main{} initialized from the backbone of Llama 3.2 3B (base).
Our \model{} uses 4 Mamba-2 layers without MLPs for both the encoder and decoder with a hidden dimension of $1536$.
Because the Llama model has a hidden dimension of $3072$, we add MLP adapters with hidden dimension $8192$ after chunking and right before dechunking (i.e. right before and after feeding into the main stage).
We train the model for 90000 gradient steps with sequence length $8192$ and batch size $256$, for a total of 189B bytes.

\paragraph{Aligning the encoder.} The primary difficulty in converting a tokenized model into a byte-level one is that the encoder and DC must produce chunks that the tokenized model can produce useful output with.
Thus, our training (besides using the standard next-byte prediction loss), adds the following losses (see \cref{fig: distill} for a visual).
\begin{enumerate}
    \item A binary cross-entropy boundary-prediction loss (with equal weight as the main loss) that operates on the \router{} probabilities and targets the router to pass \emph{the start of every real token} through the \main{}.
    \item A hidden state matching loss that matches the post-adapter hidden state with the ``correct" hidden state.
        Here, if $\hat z_k$ is the hidden representation that was passed into the \main{} at (byte) position $t$, we try to match $z_k$ with the embedding of the token that the $t$th byte was part of, \emph{except} when the $t$h byte is the first byte of its token, in which case we match the $z_t$ with the \emph{previous} token's embedding.
        Embedding matching is done with an L2 loss with a weight of 0.02.
\end{enumerate}

In the ideal case where both losses are zero, the router sends exactly the first byte of each token through to the \main{} with the right embedding.
The \main{} would thus see exactly the representation it would see with a tokenizer + embedding setup.
In practice, sending both losses to zero is literally impossible, as we discuss below.
However, we still find that the boundary-prediction loss is crucial for learning a good matching, while the embedding-matching loss is helpful in speeding up training but not necessary.
In fact, increasing the loss weight on the embedding-matching loss too much can harm the language-modeling loss.

\paragraph{Tokenization bias.} We are not able to send all auxiliary losses to zero because online prediction of BPE boundaries is an impossible task.
\citet{phan2025exact} coined the term ``tokenization bias" to represent the fact that the tokenization process implicitly contains next-byte information.
For example, the Llama 3 tokenizer tokenizes the strings \texttt{\_distill} and \texttt{\_distinct} into \texttt{[\_dist, ill]} and \texttt{[\_distinct]}.
Prior use of this term has been to suggest that if an autoregressive language model is prompted with \texttt{\_dist}, the nature of its training will be that it will never complete with \texttt{inct} (this is in fact a flaw of all tokenization-based models).

For us, however, tokenization bias implies that we cannot determine whether or not the \texttt{i} in \texttt{\_disti} is the start of a new word until \emph{after} seeing the next character.
In fact, the problem can be even worse--consider \texttt{\_applicable} (becomes \texttt{[\_app, licable]}) and \texttt{\_applicant} (becomes \texttt{[\_applicant]}): Determining whether \texttt{l} is the start of a token requires knowing the next two bytes as well.

While the \model{} does use the \main{}, it is not able to exactly match the behavior of the original tokenized model.
Instead, it is finding slightly different representations of tokens to use in the main stage. 
Recent work has shown that tokenized language models can process tokenization sequences distinct from the ``canonical" greedy tokenization \citep{vieira2024language}, so it is possible our \model{} found another alternate representation that the pretrained model could process.

\textbf{Remark.} One might ask if our distilled model has simply learned to tokenize on spaces (since spaces are always the start of a new token). It has not. Simply tokenizing on spaces would yield a sub-95\% boundary prediction accuracy; however, our distilled model gets boundary prediction accuracy above 99.5\%. This suggests that the resulting \model{} is able to recognize some, but not all, subword boundaries.


\paragraph{Results.} The results from our distillation procedure are shown in \cref{tab: distill}. 
H-Net is able to approximately match performance across almost all benchmarks; in general, H-Net is not able to replicate the behavior of the tokenized model exactly, so it is not unexpected that the benchmarks are slightly worse.
Byte-Latent Transformer \citep[Table 5]{BLT} performs a similar experiment, and they see a greater gap among most benchmarks (particularly PiQA, Arc-Easy, and Arc-Challenge) despite using a much larger amount of data (220B \emph{tokens} versus 189B \emph{bytes});
it is possible that this performance difference is due to the fact that a BLT module cannot be supervised to align boundaries the way that end-to-end DC can.

\input{fig/tokenized_positions_noisy}
